Clustering Beat-Chroma Patterns\\ in a Large Music Database

### 2010-03-25 Paper/draft_ron_tbm_edit.tex
final edit of Ron and Thierry before Dan's one

What are the common features of pop music?  There are conventions
of subject matter, instrumentation, form, rhythm, harmony, and melody, among
others.  Our interest here is in the tonal content of the music -- \ie the
harmony and melody.  As a computationally-convenient proxy for a richer
description of the tonal content of audio, we use the popular Chroma
representation, which collapses an acoustic spectrum into a 12-dimensional
description, with one bin for each semitone of the western musical octave.
In addition, we simplify the time axis of our representation to take advantage
of the strong beat present in most pop music, and record just one
chroma vector per beat.  This beat-synchronous chromagram
representation represents a typical music track
in a few thousand values, yet when resynthesized back into audio via
modulation of octave-invariant ``Shepard tones'', the melody and
chord sequences of the original music usually remain recognizable
\cite{Ellis2007a}.  To the extent, then, that beat-chroma
representations preserve tonal content, they
are an interesting domain in which to search for patterns -- rich enough
to generate musically-relevant results, but simplified enough to
abstract away aspects of the original audio such as
instrumentation and other stylistic details.

Specifically, this paper identifies common patterns in beat-synchronous
chromagrams by learning codebooks from a large set of examples.
The individual codewords consist of short beat-chroma patches of
between 1 and 8 beats, which are able to represent the entire dataset
of millions of patterns with minimum error given a small codebook of
a few hundred entries.  Our goal is to identify meaningful information
about the musical structure represented in the entire database by
examining individual entries in this codebook.  Since the common
patterns represent a higher-level description of the musical content
than the raw chroma,
we also expect them to be useful in other applications, such as
music classification and retrieving tonally-related items.

One major issue in working with small patches concerns temporal
alignment. Learning the same motifs with different offsets 
(in time or beats) is redundant. We take advantage
of rhythmic knowledge about the song to align patches with beats.
We can further group beats into bars to produces patches
for meaningful music segments.
Our use of the Echo Nest API \cite{EchoNest} for this purpose is 
explained in Section \ref{ssec:beatpatch}. Patches built around
these temporal landmarks should capture music motifs more easily.

Prior work using small patches of chroma features includes
the ``shingles'' of \cite{Casey2007}, which were used to
identify ``remixes'', i.e., music based on some of the same
underlying instrument tracks, and also for matching
performances of Mazurkas \cite{Casey2008}.  That work,
however, was not concerned with extracting the deeper
common patterns underlying different pieces (and did not
use either beat- or bar-synchronous features).  Earlier work in
beat-synchronous analysis includes \cite{Bartsch2001},
which looked for repeated patterns within single songs to
identify the chorus, and \cite{Ellis2007a}, which 
cross-correlated beat-chroma matrices to match cover
versions of pop music tracks.  None of these works
examined or interpreted the content of the chroma
matrices in any detail.  In contrast, here we hope to
develop a codebook whose entries are of interest
in their own right.

\begin{figure}[thb]
\begin{center}
\includegraphics[width=.6\columnwidth]{code}
\end{center}
\caption{\small{A typical codeword from a codebook of size $200$ (code
    7 in Figure \ref{fig:codes1}), corresponding to a
    tonic-subdominant chord progression.  The patch is composed of $2$
    bars and the pattern length was set to $8$ beats.  }}
\label{fig:code}
\end{figure}

@@

We use the bar segmentation obtained from the Echo Nest analysis to
break a song into a collection of beat-chroma ``patches'', typically
one or two bars in length.
%
Because the bar length is not guaranteed to be 4 beats, 
depending on the meter of a particular
song, we resample each patch to a fixed length of 4 beats
per bar (except where noted).  However, the majority (82\%) of our training data
consisted of bars that were 4 beats long, so this resampling 
usually had no effect.  Most of the remaining bars (10\%) were 3 beats in
length.
% although the majority of the data contains bars that are 4 beats long
The resulting patches consist of $12 \times 4$ or $12 \times 8$ matrices.

@@

The representation resulting from this process is invariant to both
the key and meter of the original song.  This enables the study of
broad harmonic patterns found throughout the data, without regard for
the specific musical context.
%
% In the context of clustering, this enables us to study harmonic
% pattern usage without maintaining clusters for the specific context a
% particular patterns occurs in,
In the context of clustering this avoids problems such as obtaining separate
clusters for every major triad for both duple and triple meters.
% Also, this implies that the pitch envelope is important rather than
% the actual pitches.
%FIXME: Something about why we're not interested in specific chords?
%
%Specific chords do not add much information about a music collection.
%A cover song can be performed in a different key, but the motifs it
%contains remain the same.
%%%% dpwe: I don't agree with this, I'd rather say nothing

\subsection{Clustering}

% We can then stack beat vectors for a number of bars (usually 1 or 2).

We use an online version of the vector quantization algorithm
\cite{Gersho1991} to cluster the beat-chroma patches described in the
previous section.
% This can also be seen as an online K-means.
For each sample from the data, the algorithm finds the closest cluster
in the codebook and updates the cluster centroid (codeword) to be
closer to the sample according to a learning rate $\ell$.
The clusters are updated as each data point is seen, as opposed to
once per iteration in the standard $k$-means algorithm.
The details are explained in Algorithm \ref{algo:vq}.
%
As in standard $k$-means clustering, the codebook is initialized by
choosing $K$ random points from our dataset.
%
Note that this algorithm, although not optimal, scales linearly with
the number of patches seen and can be interrupted at any time
to obtain an updated codebook. %not that this is important...
% The setting of the learning rate might become trickier, but this is
% not fundamentally a problem.

@@

\subsection{Settings}\label{ssec:setting}
We take one or two bars and resample the patches to 4 or 8 columns 
respectively.  We learn a codebook of size $K$ over the Cowbell
dataset using the online VQ algorithm (Algorithm \ref{algo:vq}). We use a
learning rate of $\ell=0.01$ for $200$ iterations over the whole
dataset.
%
We then use the resulting codebook to encode the test set.
Each pattern is encoded with only one code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, 
%\ie the size of the cluster around a code.
i.e., the proportion of patterns that quantize to each code.

We use the average squared Euclidean distance as the distortion
measure between chroma patches.  Given a pattern $p_1$
composed of elements $p_1(i,j)$, and a similar pattern $p_2$, the
distance between them is:
\begin{eqnarray}
  \dist(p_1,p_2) = \sum_{i,j} \frac{(p_1(i,j) - p_2(i,j))^2}{\size(p_1)}
  \label{eq:dist}
\end{eqnarray}
We assume $p_1$ and $p_2$ have the same size.  This is enforced by the
resampling procedure described in Section \ref{sec:approach}.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.8\columnwidth]{data_sizes}
\end{center}
\caption{\small{Distortion for a codebook of size $100$ encoding one bar
at a time with by $4$ columns.
Therefore, each codeword has $12 \times 4 = 48$ elements.
Distortion is measurde on the test set.  Training data sizes range
from $0$ (just initialization) to $500,000$. Patterns were selected at
random from the dataset of approximately $3.7$ million patterns.
}}
\label{fig:data_sizes}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{codes1}
\end{center}
\caption{\small{
The 25 codes that are 
most commonly used for the {\it artist20} test set. 
Codes are from the $200$-entry codebook trained on $2$ bar, $12 \times 8$
patches.  The proportion of patches accounted for by each pattern is shown 
in parentheses.
%There are $71832$ patterns in total in the dataset.
}}
\label{fig:codes1}
\end{figure}

\begin{figure}[bt]
\begin{center}
\includegraphics[width=.99\columnwidth]{freqs}
\end{center}
\caption{\small{Usage proportions for all 200 codewords 
on the {\it artist20} test set (which comprises $71,832$
patterns).  Also shown are the usage proportions for the training set (``cowbell''), 
which are similar.
Note that even though all codewords are initialized from samples, some
are used only once in the training set, or not at all for test set.
}}
\label{fig:freqs}
\end{figure}

The 25 most frequently used codewords in the test set are shown in
Figure \ref{fig:codes1}.  
The frequency of use of these codewords is shown in Figure \ref{fig:freqs}.
The codewords primarily consist of sustained
notes and simple chords.  Since they are designed to be key-invariant,
specific notes do not appear.  Instead the first 7 codewords
correspond to a single note sustained across two bars (codeword 0),
perfect fifth (codewords 1 and 2) and fourth intervals (codewords 3
and 6, noting that the fourth occurs when the per-pattern transposition 
detects the fifth rather than the root as the strongest chroma bin, and vice-versa), and a
major triads transposed to the root and fifth (codewords 5 and 4, respectively).  Many of the remaining
codewords correspond to common transitions from one chord to another,
\eg a V-I transition in codes 7 and 9
(e.g., Gmaj
$\rightarrow$ Cmaj, or G5 $\rightarrow$ C5 as a guitar power chord)
% assuming that chroma 0 is C) 
and the reverse I-V transition in code 21 (e.g., Cmaj
$\rightarrow$ Gmaj).
%FIXME: Does this transition have a better name like ``relative x''?
% C5 -> F5 in 'guitar chords', basically a power chord

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes_lle}
\end{center}
\caption{\small{
LLE visualization of the codebook.
%LLE tries to keep near neighbors (here according to euclidean distance)
%in the same $2$D neighborhood. 
Shown patterns are randomly selected
from each neighborhood.
}}
\label{fig:lle}
\end{figure}

In an effort to visualize the span of the entire codebook, we used 
Locally linear embedding (LLE) \cite{Roweis2000}\footnote{implementation:
  \url{http://www.astro.washington.edu/users/vanderplas/coding/LLE/}}
to arrange the codewords on a 2D plane while keeping similar 
patterns as neighbors.  Figure \ref{fig:lle} shows the resulting 
distribution along with a sampling of patterns; notice trends such as the 
more ``wideband'' patterns to the right.

%Another interesting visualization uses LLE\footnote{implementation:
%  \url{http://www.astro.washington.edu/users/vanderplas/coding/LLE/}}
%algorithm \cite{Roweis2000} to display codewords arranged in
%coherent groups. See Figure \ref{fig:lle}. For instance, we see
%``wideband'' patterns grouping to the right.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{code_variance}
\end{center}
\caption{\small{Average variance of codewords along the time dimension. The vertical
axis cuts at the $53$rd pattern, roughly the number of 
codewords consisting entirely of sustained chords.  Representative patterns are shown 
in each range.
}}
\label{fig:code_var}
\end{figure}

Noting that many of the codewords reflect sustained patterns with 
little temporal variation, Figure \ref{fig:code_var} plots the average 
variance along time of all 200 patterns.  
Some 26\% of the codewords have very low variance, corresponding 
to stationary patterns similar to the top row of Figure
\ref{fig:codes1}. 
%For $2$ bars codewords, it seems that patterns
%containing more complex temporal activity account for only $33$\% of
%the data.
%%%% dpwe: I can't tell what this means, so drop it.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{codes_8bars}
\end{center}
\caption{\small{Sample of longer codewords spanning $8$
    bars. Codewords were randomly selected from a 100-entry 
    codebook. Percentages of use are shown in parentheses.  Most patterns
    consist of sustained notes or chords, but code 0 shows one-bar 
    alternations between two chords, and code 4 contains two cycles of a
    1$\rightarrow$1$\rightarrow$1$\rightarrow$2 progression.}}
\label{fig:codes8bars}
\end{figure}

We made some preliminary experiments with codebooks based on 
longer patches.  Figure \ref{fig:codes8bars} presents a codewords 
from an 8 bar (32 beat) codebook.  We show a random selection 
since all the most-common codewords were less interesting, sustained 
patterns.

%Showing codewords of $2$ bars is practical and consistent with our experiments
%in Section \ref{sec:exps2}. However, longer codewords can also reveal
%interesting structure. A random sample of such codewords is presented
%in Figure \ref{fig:codes8bars}. Most used codewords would be less interesting,
%as they all are sustained patterns.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{close_patterns1}
\end{center}
\caption{\small{Cluster around centroid presented in
Figure \ref{fig:code}. Taken from the {\it artist20} dataset, the cluster
size is actually $639$. Shown samples were randomly selected.
}}
\label{fig:cluster}
\end{figure}

\subsection{Within-cluster behavior}
In addition to inspecting the codewords, it is important to understand 
the nature of the cluster of patterns represented by each codeword, i.e., 
how well the centroid describes them, and the kind of detail that has 
been left out of the codebook.  Figure \ref{fig:cluster} shows a random 
selection of the $639$ patterns from the {\it artist20} test set that 
were quantized to codeword 7 from Figure \ref{fig:codes1}, the V-I cadence. 
Figure \ref{fig:cluster_diff} illustrates the difference between the actual 
patterns and the quantized codeword for the first three patterns; although 
there are substantial differences, they are largely unstructured, indicating 
that the codebook has captured the important underlying trend.

%Visualizing a codebook is half of the story, we also want to understand
%clusters, \ie the patterns from actual song that
%surround a given centroid (codeword). Unsurprisingly, real patterns are
%noisier than codes. However, centroids seem to capture important aspects
%of the patterns. We could compare the centroids with the result of
%a lowpass filtering. In Figure \ref{fig:cluster} we show elements of the same
%cluster, and in Figure \ref{fig:cluster_diff}, we emphasize the difference
%with the centroid. We hope for the difference to look uniformly random,
%the result seems to fit that assumption.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.8\columnwidth]{close_patterns_diff}
\end{center}
\caption{\small{First three patterns of figure \ref{fig:cluster}
($2$nd line) presented with the centroid from Figure \ref{fig:code}
($1$st line) and the absolute difference between both ($3$rd line).
}}
\label{fig:cluster_diff}
\end{figure}

@@

Figure \ref{fig:encodesong} gives an example of encoding a song
using the codebook, showing both the full, original data, and the 
reconstruction using only the quantized codewords (at the correct transpositions).  
The quantized representation retains
the essential harmonic structure of the original features, but has smoothed
away much of the detail below the level of the 2 bar codewords.
%
%This is consistent with the codebook described in Section
%\ref{sec:codebook} which primarily consists of sustained chords across
%4 or 8 beats.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{song_encoded}
\end{center}
\caption{\small{Good Day Sunshine by \textit{The Beatles}.
Original song and encoding with a $200$ entry codebook of 
$2$ bar patterns.
}}
\label{fig:encodesong}
\end{figure}

\section{Applications}\label{sec:exps2}
We present two applications of the beat-chroma codebooks
to illustrate how the ``natural'' structure identified via 
unsupervised clustering can provide useful 
features for subsequent supervised tasks.  We will discuss how 
the codewords can be used in song segmentation, and for artist
recognition.  
%The representation created from large collections
%captures meaningful information without being forced to do so.
%
% Although our algorithm was not devised for any of these
% tasks, but some information contained in the centroids can be
% naturally used.

@@

Since the clustering described in Section \ref{sec:approach} is based
on the segmentation of the signal in to bars, 
the codewords should contain information related to bar
alignment, such as the presence of a strong beat on the first beat.
%
In this section we investigate using the codebook to identify the
bar segmentation of new songs.  
% using the information contained in the cluster centroids.  
We train a codebook of size $100$ on bars resampled to $4$ beats. Then,
we take the longest sequence of bars of $4$ beats for each song
in the test set (to avoid the alignment skew that results from spanning 
irregularly-sized bars).  
%As we mentioned before, $4$ beats is the most common bar size using the
%Echo Nest segmentation. 
We then encode each of these sequences using an offset
of $0$, $1$, $2$ or $3$ beats, and record for each song 
the offset giving the lowest distortion.  
%The best offset for a test sequence
%(the one with the lowest distortion)
%should correspond to the bar alignment in the original
%feature matrix.  
%The correct answer is an offset of $0$.
%
The results in Table \ref{tab:offset}
show that the ``true'' offset of $0$ beats is chosen in $62\%$ of cases (where a random 
guess would yield $25\%$).  Thus, the codebook is useful for identifying bar (downbeat) 
alignments.  A more flexible implementation of this idea would use dynamic 
programming to align bar-length patterns to the entire piece, including the 
possibility of 3- or 5-beat bars (as are sometimes needed to accommodate 
beat tracking errors) with an appropriate associated penalty.

@@

%The result presented here suggest a dynamic programming setting where
%we could find the optimal bar segmentation according to our codebook. 
%We could try
%all possible bar length (in terms of beats) and not restrict ourselves
%to $4$ beats bars and offsets between $0$ and $3$. This is part of
%our future work.

\begin{table}
\begin{center}
\begin{tabular}{cc}
\toprule
Offset & \% of times chosen \\
\midrule
0 & $\mathbf{62.6}$\\
1 & $16.5$\\
2 & $\text{ }9.4$\\
3 & $11.5$\\
\bottomrule
\end{tabular}
\end{center}
\caption{\small{
Song segmentation experiment:
offsets relative to ground-truth 4-beat bar boundaries 
that gave minimum distortion encodings from the bar-aligned codebook.
%The real one is $0$.
%Random guess gives an accuracy of $25$\%.
}}
\label{tab:offset}
\end{table}

@@

We apply our codebook to a simple artist recognition task.
We use the {\it artist20} data set, composed of $1402$ songs from $20$ artists, 
mostly rock and pop of different subgenres.  
Previously published results using GMMs on MFCC features achieve an 
accuracy of $59\%$, whereas using only chroma as a representation yields an
accuracy of $33\%$  \cite{Ellis2007}.

Although beat-chroma quantization naturally discards information 
that could be useful in artist classification, it is interesting to 
investigate whether some artist use certain patterns more often than others.

The dataset is encoded as histograms of the codewords 
used for each song, with frequency values normalized by the number
of patterns in the song. We test each song in a leave-one-out setting, 
and represent each of the $20$ artists by the average of their (remaining) 
song-histograms.  The test song is matched to an artist based on 
minimum Euclidean distance to these per-artist averages.
This gives an accuracy of $23.4\%$, where the random baseline 
is around $5\%$. 
The confusion matrix can be seen in Figure \ref{fig:conf_mat}, showing 
that certain artists are recognized at an accuracy far above the average.

\begin{figure}[t]
%\begin{center}
\includegraphics[width=.9\columnwidth]{conf_mat_per_artist}
%\end{center}
\caption{\small{Confusion matrix for the artist recognition task.}}
\label{fig:conf_mat}
\end{figure}

>>>>

What are the common features of pop music?  There are conventions
of subject matter, instrumentation, form, rhythm, harmony, and melody, among
others.  Our interest here is in the tonal content of the music -- \ie the
harmony and melody.  As a computationally-convenient proxy for a richer
description of the tonal content of audio, we use the popular Chroma
representation, which collapses an acoustic spectrum into a 12-dimensional
description, with one bin for each semitone of the western musical octave.
In addition, we simplify the time axis of our representation to take advantage
of the strong beat present in most pop music, and record just one
chroma vector per beat.  This beat-synchronous chromagram
representation represents a typical music track
in a few thousand values, yet when resynthesized back into audio via
modulation of octave-invariant ``Shepard tones'', the melody and
chord sequences of the original music usually remain recognizable
\cite{Ellis2007a_small}.  To the extent, then, that beat-chroma
representations preserve tonal content, they
are an interesting domain in which to search for patterns -- rich enough
to generate musically-relevant results, but simplified enough to
abstract away aspects of the original audio such as
instrumentation and other stylistic details.

Specifically, this paper identifies common patterns in beat-synchronous
chromagrams by learning codebooks from a large set of examples.
The individual codewords consist of short beat-chroma patches of
between 1 and 8 beats, optionally aligned to bar boundaries.
%
The additional temporal alignment eliminates redundancy that would be
created by learning multiple codewords to represent the same motive at
multiple beat offsets.
%
The codewords are able to represent the entire dataset
of millions of patterns with minimum error given a small codebook of
a few hundred entries.
%
Our goal is to identify meaningful information
about the musical structure represented in the entire database by
examining individual entries in this codebook.  Since the common
patterns represent a higher-level description of the musical content
than the raw chroma,
we also expect them to be useful in other applications, such as
music classification and retrieving tonally-related items.

Prior work using small patches of chroma features includes
the ``shingles'' of \cite{Casey2007_small}, which were used to
identify ``remixes'', i.e., music based on some of the same
underlying instrument tracks, and also for matching
performances of Mazurkas \cite{Casey2008}.  That work,
however, was not concerned with extracting the deeper
common patterns underlying different pieces (and did not
use either beat- or bar-synchronous features).  Earlier work in
beat-synchronous analysis includes \cite{Bartsch2001_small},
which looked for repeated patterns within single songs to
identify the chorus, and \cite{Ellis2007a_small}, which 
cross-correlated beat-chroma matrices to match cover
versions of pop music tracks.  None of these works
examined or interpreted the content of the chroma
matrices in any detail.  In contrast, here we hope to
develop a codebook whose entries are of interest
in their own right.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\columnwidth]{code}
\end{center}
\caption{\small{A typical codeword from a codebook of size $200$ (code
    7 in Figure \ref{fig:codes1}), corresponding to a
    tonic-subdominant chord progression.  The patch is composed of $2$
    bars and the pattern length was set to $8$ beats.  }}
\label{fig:code}
\end{figure}

@@

We use the bar segmentation obtained from the Echo Nest analysis to
break a song into a collection of beat-chroma ``patches'', typically
one or two bars in length.
%
Because the bar length is not guaranteed to be 4 beats, 
depending on the meter of a particular
song, we resample each patch to a fixed length of typically 4 beats
per bar.  However, the majority (82\%) of our training data
consisted of bars that were 4 beats long, so this resampling 
usually had no effect.  Most of the remaining bars (10\%) were 3 beats in
length.
% although the majority of the data contains bars that are 4 beats long
The resulting patches consist of $12 \times 4$ or $12 \times 8$ matrices.

@@

The representation resulting from this process is invariant to both
the key and meter of the original song.  This enables the study of
broad harmonic patterns found throughout the data, without regard for
the specific musical context.
%
% In the context of clustering, this enables us to study harmonic
% pattern usage without maintaining clusters for the specific context a
% particular patterns occurs in,
In the context of clustering this avoids problems such as obtaining separate
clusters for every major triad in duple or triple meter.
% Also, this implies that the pitch envelope is important rather than
% the actual pitches.
%FIXME: Something about why we're not interested in specific chords?
%
%Specific chords do not add much information about a music collection.
%A cover song can be performed in a different key, but the motifs it
%contains remain the same.
%%%% dpwe: I don't agree with this, I'd rather say nothing

\subsection{Clustering}
% We can then stack beat vectors for a number of bars (usually 1 or 2).

We use an online version of the vector quantization algorithm
\cite{Gersho1991} to cluster the beat-chroma patches described in the
previous section.
% This can also be seen as an online K-means.
For each sample from the data, the algorithm finds the closest cluster
in the codebook and updates the cluster centroid (codeword) to be
closer to the sample according to a learning rate $\ell$.
The clusters are updated as each data point is seen, as opposed to
once per iteration in the standard $k$-means algorithm.
The details are explained in Algorithm \ref{algo:vq}.
%
As in standard $k$-means clustering, the codebook is initialized by
choosing $K$ random points from our dataset.
% ADDED BY TBM AFTER DAN
Initialization had no significant effect on the distortion
in our experiments.
%
Note that this algorithm, although not optimal, scales linearly with
the number of patches seen and can be interrupted at any time
to obtain an updated codebook. %not that this is important...
% The setting of the learning rate might become trickier, but this is
% not fundamentally a problem.

@@

\subsection{Setting}\label{ssec:setting}
We take one or two bars and resample the patches to 4 or 8 columns 
respectively.  We learn a codebook of size $K$ over the Cowbell
dataset using the online VQ algorithm (Algorithm \ref{algo:vq}). We use a
learning rate of $\ell=0.01$ for $200$ iterations over the whole
dataset.
%
We then use the resulting codebook to encode the test set.
Each pattern is encoded with only one code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, 
%\ie the size of the cluster around a code.
i.e., the proportion of patterns that quantize to each code.

We use the average squared Euclidean distance as the distortion
measure between chroma patches.  Given a pattern $p_1$
composed of elements $p_1(i,j)$, and a similar pattern $p_2$, the
distance between them is:
\begin{eqnarray}
  \dist(p_1,p_2) = \sum_{i,j} \frac{\left(p_1(i,j) - p_2(i,j)\right)^2}{\size(p_1)}
  \label{eq:dist}
\end{eqnarray}
We assume $p_1$ and $p_2$ have the same size.  This is enforced by the
resampling procedure described in Section \ref{sec:approach}.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.8\columnwidth]{data_sizes}
\end{center}
\caption{\small{Distortion for a codebook of size $100$ encoding one bar
at a time with by $4$ columns.
Therefore, each codeword has $12 \times 4 = 48$ elements.
Distortion is measured on the test set.  Training data sizes range
from $0$ (just initialization) to $500,000$. Patterns were selected at
random from the dataset of approximately $3.7$ million patterns.
}}
\label{fig:data_sizes}
\end{figure}

@@

The 25 most frequently used codewords in the test set are shown in
Figure \ref{fig:codes1}.  
The frequency of use of these codewords is shown in Figure \ref{fig:freqs}.
The codewords primarily consist of sustained
notes and simple chords.  Since they are designed to be key-invariant,
specific notes do not appear.  Instead the first 7 codewords
correspond to a single note sustained across two bars (codeword 0),
perfect fifth (codewords 1 and 2) and fourth intervals (codewords 3
and 6, noting that the fourth occurs when the per-pattern transposition 
detects the fifth rather than the root as the strongest chroma bin, and vice-versa), and a
major triads transposed to the root and fifth (codewords 5 and 4, respectively).  Many of the remaining
codewords correspond to common transitions from one chord to another,
\eg a V-I transition in codes 7 and 9
(e.g., Gmaj
$\rightarrow$ Cmaj or G5 $\rightarrow$ C5 as a guitar power chord)
% assuming that chroma 0 is C) 
and the reverse I-V transition in code 21 (e.g., Cmaj
$\rightarrow$ Gmaj).
%FIXME: Does this transition have a better name like ``relative x''?
% C5 -> F5 in 'guitar chords', basically a power chord

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes1}
\end{center}
\caption{\small{
The 25 codes that are 
most commonly used for the {\it artist20} test set. 
Codes are from the $200$-entry codebook trained on $2$ bar, $12 \times 8$
patches.  The proportion of patches accounted for by each pattern is shown 
in parentheses.
%There are $71832$ patterns in total in the dataset.
}}
\label{fig:codes1}
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.99\columnwidth]{freqs}
\end{center}
\caption{\small{Usage proportions for all 200 codewords 
on the {\it artist20} test set (which comprises $71,832$
patterns).  Also shown are the usage proportions for the training set (``cowbell''), 
which are similar.
Note that even though all codewords are initialized from samples, some
are used only once in the training set, or not at all for test set.
}}
\label{fig:freqs}
\end{figure}

An alternate visualization of the codebook is shown in Figure
\ref{fig:lle}.  We use locally linear embedding \cite{Roweis2000} to
group near neighbors on a 2 dimensional plane\footnote{We used the
  implementation from:
  \url{http://www.astro.washington.edu/users/vanderplas/coding/LLE/}}.
This visualization highlights different groups of codewords: sustained
chords on the top left, chord changes on the bottom left, and more
complex sustained chords and ``wideband'' noisy patterns grouping to
the right of the figure.

Given the prevalence of simple codewords in Figures \ref{fig:codes1}
and \ref{fig:lle}, we assess the proportion of the codebook devoted to
representing sustained chord patterns.  The average variance over time
within each codeword is plotted in Figure \ref{fig:code_var}.  From
this figure it is clear that 25\% of the codebook corresponds to very
simple sustained patterns similar to the top row of Figure
\ref{fig:codes1}. For $2$ bar codewords, it seems that patterns
containing more complex temporal activity account for only $33$\% of
the data.

Showing codewords of $2$ bars is practical and consistent with our experiments
in Section \ref{sec:exps2}. However, longer codewords can also reveal
interesting structure. A random sample of such codewords is presented
in Figure \ref{fig:codes8bars}. Most used codewords would be less interesting,
as they all are sustained patterns.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{codes_lle}
\end{center}
\caption{\small{ Locally linear embedding (LLE) visualization of the
    codebook described in Section \ref{sec:visu}.  Each codeword is
    represented by a point on the scatter plot.  Randomly selected
    codewords from the embedded space are displayed over the plot.
    %FIXME: does this figure really provide any new insight?
  }}
\label{fig:lle}
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{code_variance}
\end{center}
\caption{\small{Average variance in time of a codebook. The vertical
axis cuts at the $63$rd pattern, a rough estimate of the number of 
codewords consisting of sustained chords.
}}
\label{fig:code_var}
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.8\columnwidth]{codes_8bars}
\end{center}
\caption{\small{Sample of longer codewords spanning over $8$
    bars. Codewords selected randomly from the codewords used in the
    test set. Percentage of use display in parentheses.  Most patterns
    consist of sustained note or chords, but code 0 shows a repeating
    two chord pattern and code 4 contains two bars of a
    1$\rightarrow$1$\rightarrow$1$\rightarrow$2 progression.}}
\label{fig:codes8bars}
\end{figure}

\subsection{Cluster}
Clusters are formed around a centroid (codeword) by all patterns closer
to that codeword than any other. Since the centroid is somewhat an
average of those patterns, the latter tend to be noisier. Hence, codewords
can be seen as the output of a lowpass filtering mechanism.
In Figure \ref{fig:cluster} we show elements of one given
cluster, and in Figure \ref{fig:cluster_diff}, we emphasize the difference
with the centroid. We hope for the difference to look uniformly random,
the result seems to fit that assumption.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{close_patterns1}
\end{center}
\caption{\small{Cluster around centroid presented in
Figure \ref{fig:code}. Taken from the artist20 dataset, the cluster
size is actually $639$. Shown samples were randomly selected.
}}
\label{fig:cluster}
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.8\columnwidth]{close_patterns_diff}
\end{center}
\caption{\small{First three patterns of figure \ref{fig:cluster}
($2$nd line) presented with the centroid from Figure \ref{fig:code}
($1$st line) and the absolute difference between both ($3$rd line).
}}
\label{fig:cluster_diff}
\end{figure}

@@

Figure \ref{fig:encodesong} shows an example of a song encoded
using the codebook described in Section \ref{sec:codebook}.  Both the
Echo Nest chroma features for the example song and the corresponding
quantization are displayed.  The quantized representation retains
dominant harmonic structure of the original features, but is smoothed
in time due to the use of 2 bar long codewords.
%
This is consistent with the codebook described in Section
\ref{sec:codebook} which primarily consists of sustained chords across
4 or 8 beats.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{song_encoded}
\end{center}
\caption{\small{Good Day Sunshine by \textit{The Beatles}.
Original song and its encoding using a $200$ entry codebook learning
$2$ bars at a time.
}}
\label{fig:encodesong}
\end{figure}

\section{Applications}\label{sec:exps2}
In this section we investigate the application of beat-chroma
codebooks to two tasks unrelated to clustering.
%
In the first experiment, we show that the codebooks retrain
information related to bar alignment and can therefore be used to
discover the bar boundaries for new songs.  The second experiment
shows that we can discriminate artists based on the encoding of their
songs, implying that some artists have a tendency to use different
beat-chroma patterns than others.

% Although our algorithm was not devised for any of these
% tasks, but some information contained in the centroids can be
% naturally used.

@@

Since the clustering described in Section \ref{sec:approach} is based
on the segmentation of the signal into bars, it stands to reason that
the cluster centroids should contain information related to bar
alignment, such as the presence of a strong beat on the first beat.
%
We investigate whether it is possible to recover the
bar segmentation of new songs using the information contained in the
cluster centroids.  The results are shown in Table \ref{tab:offset}. We find
the right segment with an accuracy of $62\%$ where random is
$25\%$. Details of the experiment follow.

We train a codebook of size $100$ on bars resized to $4$ beats. Then,
we take the longest sequence of bars of $4$ beats for each song
in the test set.  As
we mentioned before, $4$ beats is the most common bar size using the
Echo Nest segmentation. We then encode each of these sequences using an offset
of $0$, $1$, $2$ or $3$ beats. The offset on a test sequence
with the lowest distortion is our predicted alignment.
The correct answer is an offset of $0$.

@@

The result presented here suggest a dynamic programming setting where
we could find the optimal bar segmentation according to our codebook. 
We could try
all possible bar length (in terms of beats) and not restrict ourselves
to bars of $4$ beats and offsets between $0$ and $3$. This is part of
our future work.

\begin{table}[t]
\begin{center}
\begin{tabular}{cc}
\toprule
Offset & \% of times chosen \\
\midrule
0 & $\mathbf{62.6}$\\
1 & $16.5$\\
2 & $09.4$\\
3 & $11.5$\\
\bottomrule
\end{tabular}
\end{center}
\caption{\small{
Results from the song segmentation experiment.
The correct offset is $0$.
Baseline performance by choosing the offset randomly gives an accuracy of $25$\%.
}}
\label{tab:offset}
\end{table}

@@

We apply our codebook to the artist recognition task.
Our dataset, ``artist20'', encompass $1402$ songs from $20$ artists, 
mostly rock and pop of different subgenres.
The best accuracy on the dataset in the literature is $59\%$, whereas
using only chroma and a Gaussian mixture model yields an accuracy of
$33\%$ \cite{Ellis2007}.

The clustering method proposed in this work is not
intended to perform well at artist recognition. 
In particular,
we lose all timbral information which would be useful. 
However, we can
imagine that some artist use more often some patterns than others.
Our accuracy result is $23.4\%$ while random
is around $5\%$. 
The confusion matrix can be seen in Figure \ref{fig:conf_mat}.
Details of the experiment follow.

All songs in the artist dataset are encoded as an histogram of the codes
used in that song. The frequency values are normalized by the number
of patterns in the song. We test each song in a leave-one-out setting.
Leaving aside the test song, we represent each of the $20$ artists by the 
average of
the song histograms of that artist. The test song is compared to
the histogram of each artist using Euclidean distance.

\begin{figure}[t]
%\begin{center}
\includegraphics[width=.9\columnwidth]{conf_mat_per_artist}
%\end{center}
\caption{\small{Confusion matrix for artist recognition task.}}
\label{fig:conf_mat}
\end{figure}

Looking at the most typical patterns per artist can be insightful.
To find these patterns, we compare its use by one artist and divide
by its use by all artists. In Figure \ref{fig:typicalpat} we see 
Metallica's pattern, then Tori Amos and
Suzanne Vega's pattern. These three artists were easily identified.
Artists such as Metallica are characterized by ``wideband'' patterns,
containing features muddled by the noisy acoustics containing many
transients,
distortion, etc, which can be seen in Figure \ref{fig:metallica2}.

% ignoring the two following figures
\iffalse
\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\columnwidth]{metallica_pattern}
\end{center}
\caption{\small{Most typical pattern for Metallica.
COULD BE MIXED WITH TORI AMOS FIGURE.
}}
\label{fig:metallica}
\end{figure}
